State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-06-11 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
edits to the results section of the paper

%The method is appropriate for state estimation of stochastic nonlinear dynamical systems. 
\section{Results}\label{ResultsSection} The non-reduced neural field model derived in Section~\ref{NeuralModelSection} was used to generate the data for state and parameter estimation. All parameters for the model are given in Table~\ref{tab:Model Parameters}. Figure~\ref{fig:experimental design}(b) shows example data from the sensors of the model, and figure~\ref{fig:experimental design}(c) shows the power spectral density (PSD) of the simulated time-series. The PSD shows the typical $1/f$ characteristics of intracranial EEG.
\begin{table*}
\centering
\begin{tabular}{c|c|c}
	\hline\hline Symbol & Value & Units \\
	\hline\hline
	$\zeta$ & 0.01 & ms$^{-1}$\\
	$f_{max}$ & 10 & spikes/s \\
	$\varsigma$ & 0.8 & spike/mV\\
	$v_0$ & 2 & mV\\
	$\boldsymbol{\theta}$ & $\left[\begin{array}{ccc}
	10 &-8 &0.5
	\end{array}
	\right]^{\top}$ & synaptic strength\\
	$\sigma_{\psi,1}$ & 1.8 & inner connectivity kernel width\\
	$\sigma_{\psi,2}$ & 2.4 & middle connectivity kernel width\\
	$\sigma_{\psi,3}$ & 6 & outer connectivity kernel width\\
	$\Sigma$ & & field disturbance covariance \\
	sensor parameters & & \\
	$\Delta$ & 0.5 & space step \\
	$T_s$ & 1 & ms time step \\ 
	\hline\hline
\end{tabular}\label{tab:Model Parameters}
\caption{Parameters for neural field model used to generate data to demonstrate the estimation framework.}
\end{table*}
The sensors were spaced in a regular square 14 $\times$ 14 grid, the sensor width $\sigma^2_m$ was 0.81 which equals $1.5$~mm full width at half maximum, with spacing $\Delta_y$ equal to $1.5$~mm. This arrangement modelled the recording as having some cross talk, which is typical from neurophysiological recordings. %Note, sensor design that is sufficient to guarantee an accurate representation of the field can be achieved using the frequency analysis from Section~\ref{SpectralAnalysisSection} given access to the true field. 
\begin{figure}[th]\label{fig:experimental design}
   	\begin{center}
    		\subfigure[][]{\includegraphics{./Graph/ExperimentFigureEEGPy.pdf}} \\
   		\subfigure[][]{\includegraphics{./Graph/ExperimentFigurePSDPy.pdf}} \\
   	\end{center}
   	\caption{Example of experimental design and model output.  \textbf{a} Example of data generated by the non-reduced model through a subset of the observations. The sensors were spaced in a regular square $14 \times 14$ grid. \textbf{b}. Power spectral density of the data generated by an observation showing the typical $1/f$ characteristic of iEEG.} 
   \end{figure}

@@

\caption{The absolute error between true and estimated parameters averaged over 100 runs. All parameters converge after six iterations. The errors fall below $10^{-4}$ after six iterations. }

@@

Using the model selection techniques described in Section~\ref{SpectralAnalysisSection}, the spatial frequency of the field was used to confirm that the spacing and width of the sensors was adequate to capture the significant dynamics of the field. Figure~\ref{fig:FieldFFT} shows the spatial frequency of the neural field. The cut-off frequency, $\nu_c$, was taken to be 0.24~Hz ($-3$~dB point). From equation~\ref{eq:MinimumSensorDistance}, this yielded a minimum sensor separation of 2.05~mm. This confirms our separation, $\Delta_{y}$ of 1.5~mm was sufficient to prevent problems associated with spatial aliasing.
The spatial frequency from the observations is shown in Figure~\ref{fig:ObservationFreqAnal}. The cut-off frequency, $\nu_{cy}$, was taken to be $0.19$~Hz ($-3$~dB point). The observation cut-off frequency is lower than the field cut-off due to the low-pass filtering effect of the sensors. The spatial frequency response of the sensors can be seen in figure~\ref{fig:SensorFunctionFreqResponse}. Using an oversampling parameter, $\rho_{\phi}$, of $1$, the minimum distance between basis functions was calculated to be $2.63$ to satisfy Shannon's sampling criterion (from equation~\ref{eq:BasisFunctionSeparation}). From equation~\ref{eq:WidthCutOffRelationship} the required basis function width to represent the dominant spatial frequencies was calculated to be $1$. Following this, the distance between basis functions, $\Delta_{\phi}$, was set to $2.5$, which is sufficient to capture the significant dynamics according to our criteria. The basis function width, $\sigma_{\phi}^2$, was set above the calculated width to $2.5$ as a design choice. This was a result of a trade-off that ensured that the lower spatial frequency components of the field were well represented at the expense of higher spatial frequency components. The spatial frequency response of the basis function can be seen in Figure~\ref{fig:BasisFunctionFreqResponse}.
	\caption{Spatial frequency analysis. \textbf{a.} Two 2-dimensional spatial frequency (in power dB, averaged over 400 time steps) of the observed field.  project. The symmetry of the field basis functions, connectivity kernel basis functions and sensors allow for the 1-dimensional representation of the spectral characteristics. The dashed line shows the cut-off frequency (-3 dB point). \textbf{b} The spatial frequency response of a 1-dimensional model sensor. The dashed line shows the cut-off frequency (-3 dB point), indicating a wider bandwidth than the field basis function. \textbf{c.} The spatial frequency response of a 1-dimensional neural field basis function.} 
To demonstrate the performance of the parameter and state estimation, $100$ realisations of the data were generated. Each realisation consisted of 500~ms of data and the estimation was applied to the final 400~ms, allowing the model to stabilise.\dean{transient!} The initial state and parameters were unknown to the estimator. The distribution of the parameters and the resulting kernel are shown in figure~\ref{fig:Parameters} and figure~\ref{fig:KernelEstimates}, respectively. Although a wide distribution can be seen over the parameter estimates, the ratio between them remains within a narrow range as shown in figure~\ref{fig:ParametersRatio}. The state estimates for one realisation were evaluated by comparing the field reconstruction to the true field using the mean over time of the Root Mean Square Error (RMSE) and standard deviation of the RMSE giving $0.5\pm 0.02$. The RMSE of the field estimates over time for 100 realizations is also shown in figure~\ref{fig:RMSE}. A snapshot of the true field, the estimated field and their corresponding error is illustrated in figure~\ref{fig:FieldEstimate}. The convergence of the parameter estimation is shown in figure~\ref{fig:ParametersConvergence}.

@@

@@

  \begin{figure}
   	\begin{center}
   		\includegraphics{./Graph/MRMSE.pdf} 
   	\end{center}
   	\caption{The mean of average RMSE of the estimated field over 100 realisations (solid line) and the 95~\% confidence interval (grey region). The error remains consitent over time. \dean{David suggested for us to use percentage error, and use a hard threshold of points close to zero to stop problems with divide by zeros. Also, it might be good to show the data points along the line, so we can see the number of iterations it takes to stabilise.}} 
\label{fig:RMSE}
   \end{figure}
\begin{figure*}[!th]
\centering 
\subfigure[][]{\label{fig:TrueField}\includegraphics[scale=1]{./Graph/TrueField90.pdf}}
\subfigure[][]{\includegraphics[scale=1]{./Graph/EstimatedField90.pdf}}
\subfigure[][]{\includegraphics[scale=1]{./Graph/EstimationError90.pdf}}\\
\caption{Examples of the neural fields. \textbf{a.} The true field from the non-reduced model which was used to generate data. \textbf{b.} The reconstructed estimated field of the reduced model, showing the effect of the basis function decomposition on the frequency content of the estimated field. \textbf{c.} The absolute error between the true and reconstructed fields, the error shows high frequency behaviour.}
\label{fig:FieldEstimate}
\end{figure*}

>>>>

%The method is appropriate for state estimation of stochastic nonlinear dynamical systems. 
\section{Results}\label{ResultsSection} The non-reduced neural field model derived in Section~\ref{NeuralModelSection} was used to generate the data for state and parameter estimation. All parameters for the model are given in Table~\ref{tab:Model Parameters}. Figure~\ref{fig:experimental design}(b) shows example data from the sensors of the model, and figure~\ref{fig:experimental design}(c) shows the power spectral density (PSD) of the simulated time-series. The PSD shows the typical $1/f^2$ characteristics of intracranial EEG.
\begin{table*}
\begin{tabular}{c|c|c}
	\hline\hline Symbol & Value & Units \\
	\hline\hline
	$\zeta$ & 0.01 & ms$^{-1}$\\
	$f_{max}$ & 10 & spikes/s \\
	$\varsigma$ & 0.8 & spike/mV\\
	$v_0$ & 2 & mV\\
	$\boldsymbol{\theta}$ & $\left[\begin{array}{ccc}
	10 &-8 &0.5
	\right]^{\top}$ & synaptic strength\\
	$\sigma_{\psi,1}$ & 1.8 & inner connectivity kernel width\\
	$\sigma_{\psi,2}$ & 2.4 & middle connectivity kernel width\\
	$\sigma_{\psi,3}$ & 6 & outer connectivity kernel width\\
	$\Sigma$ & & field disturbance covariance \\
	sensor parameters & & \\
	$\Delta$ & 0.5 & space step \\
	$T_s$ & 1 & ms time step \\ 
	\hline\hline
\end{tabular}\label{tab:Model Parameters}
\caption{Parameters for neural field model used to generate data to demonstrate the estimation framework.}
\end{table*}
The sensors were spaced in a regular square 14 $\times$ 14 grid, the sensor width $\sigma^2_m$ was 0.81 which equals $1.5$~mm full width at half maximum, with spacing $\Delta_y$ equal to $1.5$~mm. This arrangement modelled the recording as having some cross talk, which is typical from neurophysiological recordings. %Note, sensor design that is sufficient to guarantee an accurate representation of the field can be achieved using the frequency analysis from Section~\ref{SpectralAnalysisSection} given access to the true field. 
\begin{figure}[th]\label{fig:experimental design}
    		\subfigure[][]{\includegraphics{./Graph/ExperimentFigureEEGPy.pdf}} \\
   		\subfigure[][]{\includegraphics{./Graph/ExperimentFigurePSDPy.pdf}} \\
   	\caption{Example of experimental design and model output.  \textbf{a} Example of data generated by the non-reduced model through a subset of the observations. \textbf{b}. Power spectral density of the data generated by an observation showing the typical $1/f^2$ characteristic of iEEG.} 
Using the model selection techniques described in Section~\ref{SpectralAnalysisSection}, the spatial frequency of the field was used to confirm that the spacing and width of the sensors was adequate to capture the significant dynamics of the field. Figure~\ref{fig:FieldFFT} shows the spatial frequency of the neural field. The cut-off frequency, $\nu_c$, was taken to be 0.24~Hz ($-3$~dB point). From equation~\ref{eq:MinimumSensorDistance}, this yielded a minimum sensor separation of 2.05~mm. This confirms the separation, $\Delta_{y}$, of 1.5~mm was sufficient to prevent problems associated with spatial aliasing.
The spatial frequency from the observations is shown in Figure~\ref{fig:ObservationFreqAnal}. The cut-off frequency, $\nu_{cy}$, was taken to be $0.19$~Hz ($-3$~dB point). The observation cut-off frequency is lower than the field cut-off due to the low-pass filtering effect of the sensors. The spatial frequency response of the sensors can be seen in Figure~\ref{fig:SensorFunctionFreqResponse}. Using an oversampling parameter, $\rho_{\phi}$, of $1$, the minimum distance between basis functions was calculated to be $2.63$ to satisfy Shannon's sampling criterion (from equation~\ref{eq:BasisFunctionSeparation}). From equation~\ref{eq:WidthCutOffRelationship} the required basis function width to represent the dominant spatial frequencies was calculated to be $1$. The distance between basis functions, $\Delta_{\phi}$, was set to $2.5$, which is sufficient to capture the significant dynamics according to our criteria. The basis function width, $\sigma_{\phi}^2$, was set above the calculated width to $2.5$, which reflected a design choice. This was a result of a trade-off that ensured that the lower spatial frequency components of the field were better represented at the expense of higher spatial frequency components. The spatial frequency response of the basis function can be seen in Figure~\ref{fig:BasisFunctionFreqResponse}.
	\caption{Spatial frequency analysis. \textbf{a.} Two 2-dimensional spatial frequency (in power dB, averaged over 400 time steps) of the observed field. The symmetry of the sensors and field basis functions allow for the 1-dimensional representation of the spectral characteristics. The dashed line shows the cut-off frequency (-3 dB point). \textbf{b.} The spatial frequency response of a 1-dimensional model sensor. The dashed line shows the cut-off frequency (-3 dB point), indicating a wider bandwidth than the field basis function. \textbf{c.} The spatial frequency response of a 1-dimensional neural field basis function.} 
To demonstrate the performance of the parameter and state estimation a Monte Carlo approach was employed where $100$ realisations of the data were generated. Each realisation consisted of 500~ms of data (sampled at $1$~kHz) and the estimation was applied to the final $400$~ms, allowing the model to stabilise and transients to die out from the initial conditions. The initial state and parameters were unknown to the estimator. The distribution of the estimated parameters is shown in Figure~\ref{fig:Parameters}. The distribution of the parameters estimates are centred about the true values indicating good results. Although the distributions are quite wide, the ratio of the estimates for each realisation falls into relatively tight distributions, as seen in Figure~\ref{fig:ParametersRatio}, This indicates the balance of excitation and inhibition can be estimated more accurately than the individual parameters. The reconstructed connectivity kernel from the estimated parameters can be seen in Figure~\ref{fig:KernelEstimates}. This Figure verifies the expected net shape of the connectivity kernel is in good accordance to the true kernel. Figure~\ref{fig:ParametersConvergence} shows the rate of convergence for parameter estimates.

@@

@@

@@

\caption{The mean of the absolute error between true and estimated parameters averaged over 100 runs. All parameters converge after six iterations. \dean{Parham: is this true? the distribution of the estimates of $\xi$ show a bias $>0.01$!} The errors fall below $10^{-4}$ after 6 iterations. }

@@

The accuracy for state estimates were evaluated by comparing the field reconstruction (see equation~\ref{DefFieldDecomp}) to the true field using the mean (over time) of the root mean square error (MRMSE) (over space) and standard deviation of the RMSE. The RMSE of the field estimates over time for 100 realisations is shown in Figure~\ref{fig:RMSE}. A snapshot of the true field, the estimated field and their corresponding error is illustrated in Figure~\ref{fig:FieldEstimate}. This figure highlights the loss of the high spatial frequency information in the basis function decomposition.
  \begin{figure}
   	\begin{center}
   		\includegraphics{./Graph/MRMSE.pdf} 
   	\end{center}
   	\caption{The mean of average RMSE of the estimated field over 100 realisations (solid line) and the 95~\% confidence interval (grey region). The error remains consistent over time.} 
\label{fig:RMSE}
   \end{figure}
\begin{figure*}[!th]
\centering 
\subfigure[][]{\label{fig:TrueField}\includegraphics[scale=1]{./Graph/TrueField90.pdf}}
\subfigure[][]{\includegraphics[scale=1]{./Graph/EstimatedField90.pdf}}
\subfigure[][]{\includegraphics[scale=1]{./Graph/EstimationError90.pdf}}\\
\caption{Examples of the neural fields. \textbf{a.} The true field from the non-reduced model which was used to generate data. \textbf{b.} The reconstructed field of the reduced model, showing the effect of the basis function decomposition on the frequency content of the estimated field. \textbf{c.} The absolute error between the true and reconstructed fields, the error shows high frequency behaviour.}
\label{fig:FieldEstimate}
\end{figure*}
